---
title: "Kriging to Predict Values on a Grid"
author: "Scott Haraburda"
date: "September 19, 2018"
output: html_document
---

```{r setup, include=FALSE}

#####Make sure you load any required packages.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


## Purpose.

To predict values on a grid, portrayed graphically with contours, also known as Gaussian process regression. This method takes collected spatially related data, usually mapped to a grid coordinates, and estimates the values between the data. This is often used in soil science and geology.

```{r krige,fig.align="center", echo=TRUE, warning = FALSE, message=FALSE}
library(lattice)
library(sp)
library(gstat)
library(RColorBrewer)
k<-read.csv("krige.csv")
coordinates(k) <-~X+Y
#
#  Plot the values
#
a<-spplot(k, 
          col.regions=bpy.colors(64),  
          key.space="right",
          main="Actual Values", 
          xlab="X",
          ylab="Y",
          scales=list(draw=T))

#
#  Calculate predicted values
#
k.cal <- k[1:259, ]
k.val <- k[260:359, ]
kb = cbind(k$X,k$Y)
min.x <- floor((bbox(kb)["x", "min"] - 0.1) * 10)/10
max.x <- ceiling((bbox(kb)["x", "max"] + 0.1) *  10)/10
min.y <- floor((bbox(kb)["y", "min"] - 0.1) *  10)/10
max.y <- ceiling((bbox(kb)["y", "max"] + 0.1) *  10)/10
cells.x <- (max.x - min.x) * 20
cells.y <- (max.y - min.y) * 20
k.raster <- SpatialGrid(GridTopology(c(min.x + 0.025, min.y+0.025),
                                     c(0.05,0.05),
                                     c(cells.x, cells.y)))
#
#  variogram model to the values in the calibration dataset
#
v <- variogram(value ~ 1, 
               loc = k.cal, 
               cutoff = 1.6)
vmf <- fit.variogram(v, 
                     vgm(12.5, "Pen", 1.2, 1.5))
k.grid <- krige(value ~ 1, 
                loc = k.cal, 
                newdata = k.raster, 
                model=vmf)
#
#  Plot predicted values
#

b<-spplot(k.grid, 
          zcol="var1.pred", 
          pretty=T, contour=T,
          col.regions=bpy.colors(64), 
          main="Predicted Values",
          xlab="X", 
          ylab="Y", 
          scales=list(draw=T))

```

## Actual Data
This is a graph of the actual 359 data (values ranging between 0 and 20) on a 5x6 grid.

<div align="center">
```{r image1,fig.align="center", echo=FALSE, warning = FALSE}
a
```
</div>

## Predicted Values
This is a contour graph of the predicted values on the 5x6 grid after kriging the actual values.

<div align="center">
```{r image2,fig.align="center", echo=FALSE, warning = FALSE}
b
```
</div>

## Comments

Kriging can be used to predict at specific points and is typically used to make a raster map by predicting at all nodes of a regular grid. When developing this prediction, one must consider regions with missing actual data as potential areas where the prediction doesn’t accurately predict values.
