---
title: "Wrangling: Replace Blank Cells with “NA”"
author: "Scott Haraburda"
date: "September 18, 2018"
output: html_document
---

```{r setup, include=FALSE}

#####Make sure you load any required packages.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


## Purpose.

To locate and remove blank cells (<b><font color="red">NULL</font></b>) from data files with vectors characterized as factors and replace with <b><font color="red">NA</font></b>.  This also replaces factor types..

```{r image1,fig.align="center", echo=TRUE, warning = FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(sjPlot)
null <- read.csv("null.csv")
x1 <- tab_df(null, title = "Before replacing null")
#
# convert factor to number
#
null$number <- as.numeric(as.character(null$number))
#
# convert factor to character
#
null$text <- as.character(null$text)
null$text[null$text=="NULL"] <- "NA"
#
# convert factor to logical
#
null$logical <- as.logical(null$logical)
x2 <- tab_df(null, title = "After replacing null")

```

## Before
This is a print of the table, center-aligned, of the data before replacing <b><font color="red">NULL</font></b> values with <b><font color="red">NA</font></b> using the tab_df function in the sjPlot package.

<div align="center">
```{r image2,fig.align="center", echo=FALSE, warning = FALSE}
x1
```
</div>

## After
This is a print of the table, center-aligned, of the data after replacing <b><font color="red">NULL</font></b> values with <b><font color="red">NA</font></b> using the tab_df function in the sjPlot package.

<div align="center">
```{r image3,fig.align="center", echo=FALSE, warning = FALSE}
x2
```
</div>

## Comments

Before an R program can look for answers, the data must be cleaned up and converted to a form that makes information accessible.  In this example, we converted data frame properties classified as factors into intended properties of numbers, characters, and logical.  This code also converted blank cells into cells containing NA. vegetables consume more overall from home.
